---
title: "ENVS 193DS Spring 2025 Final"
author: "Jose Cardoso"
date: "06/06/2025"
format:
  html:
    toc: true # use this to display a table of contents
execute:
  message: false # use this to make sure messages don't show up
  warning: false # use this to make sure warnings don't show up
---

# Set up

```{r}
#| message: false

# insert code to read in packages here
library(tidyverse)
library(janitor)
library(here)
library(flextable)

# insert code to read in your data here
sst <- read_csv("/Users/admin/Documents/github/ENVS-193DS_spring-2025_final/data/SST_update2023.csv")

occ <- read_csv("/Users/admin/Documents/github/ENVS-193DS_spring-2025_final/data/occdist.csv")
```

# Problem 1

## a. Transparent statistical methods

In part 1, the co-worker used the Pearson's r correlation test. In part 2, the co-worker used .

## b. More information needed



## c. Suggestions for rewriting



# Problem 2

## a. Cleaning and summarizing
```{r}
# cleaning and summarizing data
sst_clean <- sst |> #creating clean data frame
  clean_names() |> # capital to lowercase and spaces to underscores
  mutate(year = year(date), # creating new year and month columns
         month1 = month(date),
         month = month(month1, label = TRUE, abbr = TRUE)) |> 
  select(year, month, temp) |>  # selecting only the date and temp columns
  mutate(month = as_factor(month),
         month = fct_relevel(
           month, 
           "Jan", "Feb", "Mar", "Apr", "May", "Jun", "Jul", "Aug", "Sep", "Oct",            "Nov", "Dec")
         ) |> 
  mutate(year = as_factor(year),
         year = fct_relevel(
           year, "2018", "2019", "2020", "2021", "2022", "2023"
         )) |> 
  group_by(year,month) |> # group data by year and month
  summarize(mean_monthly_sst = mean(temp, na.rm = TRUE)) |> # calculate surface water temp mean of each month
  ungroup() |> 
  filter(year %in% c("2018", "2019", "2020", "2021", "2022", "2023"))

slice_sample(sst_clean, n = 5)

str(sst_clean)
```

# b. Visualize the data
```{r}
ggplot(data = sst_clean, # use sst_clean data frame
       aes(x = month, # month on the x-axis
           y = mean_monthly_sst, # monthly mean surface temp on the y-axis
           group = year, # group by year
           color = year))+ # color by year
  geom_point()+ #adding points
  geom_line()+ #adding lines
  scale_color_brewer()+ # color gradient in a single color
  theme_bw()+ # changing to black and white theme
  labs(x = "Month", # relabeling x-axis
       y = "Mean monthly sea surface temperature (°C)", # relabeling y-axis
       color = "Year")+ # relabeling legend title
  theme(panel.grid.major = element_blank(),# remove gridlines
        panel.grid.minor = element_blank(),
        legend.position = "inside", # place the legend inside the plot 
        legend.position.inside = c(0.15,0.7)) # place the legend to the left
```


# Problem 3

## a. Response variable

The 1s and 0s in this data set correspond to whether a nest box is or is not occupied. This was done for each species.

## b. Purpose of study

Swift Parrots are the target species for restoration efforts, while Tree Martins and Common Starlings are nontarget species that readily exploit the restoration efforts.

## c. Difference in "seasons"

The two years are years where a mast tree flowering event occurred, and the seasons are breeding events induced by the mast tree flowering event. The difference in the two years is in 2019, the second proposed restoration method was tested-permanent nesting boxes.

## d. Table of models

```{r}

```


## e. Run the models

```{r}

```


## f. Check the diagnostics

```{r}

```


## g. Select the best model

```{r}

```


## h. Visualize the model predictions

```{r}

```


## i. Write a caption for your figure



## j. Calculate model predictions

```{r}

```


## k. Interpret your results



# Problem 4

```{r}

```

